---
title: "Computational Modeling - Assignment 1"
author: "Riccardo Fusaroli"
date: "07/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rethinking)
library(tidyverse)
```

## In this assignment we learn how to assess rates from a binomial distribution, using the case of assessing your teachers' knowledge of CogSci

N.B. this markdown has 2 parts as it spans 2 weeks of teaching

### First part

You want to assess your teachers' knowledge of cognitive science. "These guys are a bunch of drama(turgist) queens, mindless philosophers, chattering communication people and Russian spies. Do they really know CogSci?", you think.

To keep things simple (your teachers should not be faced with too complicated things):
- You created a pool of equally challenging questions on CogSci
- Each question can be answered correctly or not (we don't allow partially correct answers, to make our life simpler).
- Knowledge of CogSci can be measured on a scale from 0 (negative knowledge, all answers wrong) through 0.5 (random chance) to 1 (awesome CogSci superpowers)

This is the data:
- Riccardo: 3 correct answers out of 6 questions
- Kristian: 2 correct answers out of 2 questions (then he gets bored)
- Daina: 160 correct answers out of 198 questions (Daina never gets bored)
- Mikkel: 66 correct answers out of 132 questions

Questions:

1. What's Riccardo's estimated knowledge of CogSci? What is the probability he knows more than chance (0.5) [try figuring this out. if you can't peek into chapters 3.1 and 3.2 and/or the slides]?

```{r}
name <- c("Riccardo", "Kristian", "Daina", "Mikkel")

correct <- c(3, 2, 160, 66)

total <- c(6, 2, 198, 132)

df <- cbind(name, correct, total)
```


- First implement a grid approximation (hint check paragraph 2.4.1!) with a uniform prior, calculate the posterior and plot the results
```{r}
#create grid
dens <- 1000

p_grid <- seq(from = 0 , to = 1 , length.out = dens)

prior <- rep( 1 , dens ) # Flat 


#compute likelihood
likelihood <- dbinom(3, size = 6, prob = p_grid)


#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)




#plot the posterior
df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)


ggplot(df_stats_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Riccardo")

# The probability of Riccardo knowing more than chance (0.5%)
# creating samples
n_samples <- 1e4

# make it reproducible
set.seed(3)

samples <-
  df_stats_R %>% 
  slice_sample(n = n_samples, weight_by = posterior, replace = T)

# The probability that Riccardo knows more than chance
samples %>%
  summarise(sum = mean(p_grid > .5))*100

```

- Then implement a quadratic approximation (hint check paragraph 2.4.2!).
- N.B. for the rest of the exercise just keep using the grid approximation (we'll move to quadratic approximations in two classes)

```{r}

#quadratic approximation
qa_approximation <- quap( alist(
w ~ dbinom(6,p) , # binomial likelihood
p ~ dunif(0,1) # uniform prior
),
data=list(w=3)) 

precis(qa_approximation)

curve(dnorm(x , 0.5, 0.2) , lty=2 , add=TRUE )
```


2. Estimate all the teachers' knowledge of CogSci. Who's best? Use grid approximation. Comment on the posteriors of Riccardo and Mikkel.
2a. Produce plots of the prior, and posterior for each teacher.


```{r}
#for Kristian
#compute likelihood
likelihood <- dbinom(2, size = 2, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_K <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_K, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Kristian")

p_grid[ which.max(posterior) ]


#for Daina
#compute likelihood
likelihood <- dbinom(160, size = 198, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_D <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_D, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Daina")


p_grid[ which.max(posterior) ]


#for Mikkel
#compute likelihood
likelihood <- dbinom(66, size = 132, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_M <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_M, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Mikkel")

precis(df_stats_M)
precis(df_stats_R)

```

3. Change the prior. Given your teachers have all CogSci jobs, you should start with a higher appreciation of their knowledge: the prior is a normal distribution with a mean of 0.8 and a standard deviation of 0.2. Do the results change (and if so how)?
3a. Produce plots of the prior and posterior for each teacher.

```{r}
#create grid
prior <- dnorm(p_grid, 0.8, 0.2) # SensiblyCenteredAtAssumedCogSciKnowledge
plot(p_grid, prior)

#for Riccardo
#compute likelihood
likelihood <- dbinom(3, size = 6, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Riccardo")

#for Kristian
#compute likelihood
likelihood <- dbinom(2, size = 2, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_K <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_K, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Kristian")

#for Daina
#compute likelihood
likelihood <- dbinom(160, size = 198, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_D <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_D, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Daina")


#for Mikkel
#compute likelihood
likelihood <- dbinom(66, size = 132, prob = p_grid)

#calculate posterior
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_M <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_M, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Mikkel")

```


4. You go back to your teachers and collect more data (multiply the previous numbers by 100). Calculate their knowledge with both a uniform prior and a normal prior with a mean of 0.8 and a standard deviation of 0.2. Do you still see a difference between the results? Why?

```{r}
#prior that is uniform
prior_u <- rep( 1 , dens) 
#prior that is mean adjusted
prior <- dnorm(p_grid, 0.8, 0.2) # SensiblyCenteredAtAssumedCogSciKnowledge

#for Riccardo
likelihood <- dbinom(300, size = 600, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#priors and plots
df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior_u, likelihood = likelihood)
ggplot(df_stats_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")

df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)
ggplot(df_stats_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")

#for Kristian
likelihood <- dbinom(200, size = 200, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#priors and plots
df_stats_K <- data.frame(grid = p_grid, posterior = posterior, prior = prior_u, likelihood = likelihood)
ggplot(df_stats_K, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")


df_stats_K <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)
ggplot(df_stats_K, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")
```


```{r}
#for Daina
likelihood <- dbinom(16000, size = 19800, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#priors and plots
df_stats_D <- data.frame(grid = p_grid, posterior = posterior, prior = prior_u, likelihood = likelihood)
ggplot(df_stats_D, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")

df_stats_D <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)
ggplot(df_stats_D, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")


#for Mikkel
#compute likelihood
likelihood <- dbinom(6600, size = 13200, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#priors and plots
df_stats_M <- data.frame(grid = p_grid, posterior = posterior, prior = prior_u, likelihood = likelihood)
ggplot(df_stats_M, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")


df_stats_M <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)
ggplot(df_stats_M, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability")

```


5. Imagine you're a skeptic and think your teachers do not know anything about CogSci, given the content of their classes. How would you operationalize that belief?

```{r}
# Now we expect that the teachers will perform below chance. We use Riccardo as an example 
prior <- ifelse(p_grid > 0.5 , 0 , 1) # prior that assumes below 0.5

#for Riccardo
likelihood <- dbinom(3, size = 6, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
plot(posterior)
df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

ggplot(df_stats_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/dens),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Riccardo")

```


6. Optional question: Can you estimate the difference between Riccardo's estimated knowledge and that of each of the other teachers? Would you deem it credible (that is, would you believe that it is actually different)? 

### Second part: Focusing on predictions

Last year you assessed the teachers (darned time runs quick!). Now you want to re-test them and assess whether your models are producing reliable predictions. In Methods 3 we learned how to do machine-learning style assessment of predictions (e.g. rmse on testing datasets). Bayesian stats makes things a bit more complicated. So we'll try out how that works. N.B. You can choose which prior to use for the analysis of last year's data.

Questions to be answered (but see guidance below):
1- Write a paragraph discussing how assessment of prediction performance is different in Bayesian vs. frequentist models
2- Provide at least one plot and one written line discussing prediction errors for each of the teachers.

This is the old data:
- Riccardo: 3 correct answers out of 6 questions
- Kristian: 2 correct answers out of 2 questions (then he gets bored)
- Daina: 160 correct answers out of 198 questions (Daina never gets bored)
- Mikkel: 66 correct answers out of 132 questions

This is the new data:
- Riccardo: 9 correct answers out of 10 questions (then he freaks out about teaching preparation and leaves)
```{r}
#create grid
dens <- 1000
p_grid <- seq(from = 0 , to = 1 , length.out = dens)
prior <- dnorm(p_grid, 0.8, 0.2) # SensiblyCenteredAtAssumedCogSciKnowledge

#for Riccardo - old results
#compute likelihood
likelihood <- dbinom(3, size = 6, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#Old posterior posterior
df_stats_R <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

# New results
n_success <- 9
n_trials  <- 10

# New posterior
new_R <-
  df_stats_R %>% 
  mutate(prior = posterior, 
         likelihood = dbinom(n_success, size = n_trials, prob = p_grid)) %>% 
         mutate(posterior = (likelihood * prior) / sum(likelihood*prior))


ggplot(new_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/sum(prior)),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Riccardo")
```

- Kristian: 8 correct answers out of 12 questions

```{r}
#Old data Kristian
#compute likelihood
likelihood <- dbinom(2, size = 2, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_K <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)


# New results
n_success <- 8
n_trials  <- 12

# New posterior
new_K <-
  df_stats_K %>% 
  mutate(prior = posterior, 
         likelihood = dbinom(n_success, size = n_trials, prob = p_grid)) %>% 
         mutate(posterior = (likelihood * prior) / sum(likelihood*prior))

ggplot(new_K, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/sum(prior)),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Kristian")

```


- Daina: 148 correct answers out of 172 questions (again, Daina never gets bored)
```{r}
#Old data Daina
#compute likelihood
likelihood <- dbinom(160, size = 198, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_D <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

# New results
n_success <- 148
n_trials  <- 172

# New posterior
new_D <-
  df_stats_D %>% 
  mutate(prior = posterior, 
         likelihood = dbinom(n_success, size = n_trials, prob = p_grid)) %>% 
         mutate(posterior = (likelihood * prior) / sum(likelihood*prior))

ggplot(new_D, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/sum(prior)),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Daina")

```


- Mikkel: 34 correct answers out of 65 questions
```{r}
#Old data Mikkel
#compute likelihood
likelihood <- dbinom(66, size = 132, prob = p_grid)
unstd.posterior <- likelihood * prior
posterior <- unstd.posterior / sum(unstd.posterior)

#plot the posterior
df_stats_M <- data.frame(grid = p_grid, posterior = posterior, prior = prior, likelihood = likelihood)

# New results
n_success <- 34
n_trials  <- 65

# New posterior
new_M <-
  df_stats_M %>% 
  mutate(prior = posterior, 
         likelihood = dbinom(n_success, size = n_trials, prob = p_grid)) %>% 
         mutate(posterior = (likelihood * prior) / sum(likelihood*prior))

ggplot(new_R, aes(grid,posterior)) + geom_point() +geom_line()+ geom_line(aes(grid, prior/sum(prior)),color= 'red')+ xlab("Knowledge of CogSci") + ylab("posterior probability") + ggtitle("Mikkel")
```

Guidance Tips

1. There are at least two ways of assessing predictions.
2. Last year's results are this year's expectations.
3. Are the parameter estimates changing? (way 1)
4. How does the new data look in last year's predictive posterior? (way 2)